---
title: |
  Self-Tuned Rejection Sampling within Gibbs and a Case Study in Small Area
  Estimation
subtitle: Guide to Replication Materials
author: Andrew M. Raim$^*$, Kyle M. Irimata$^*$, and James A. Livsey$^\dagger$
date: now
date-format: "YYYY-MM-DD"
thanks: |
  $^*$Center for Statistical Research and Methodology, U.S. Census Bureau.
  $^\dagger$Department of Biostatistics, University of Buffalo.
  `andrew.raim@census.gov`, `kyle.m.irimata@census.gov`, `jamesliv@buffalo.edu`.
  Document compiled `{r} format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.
format:
  pdf:
    fontsize: 10pt
    indent: false
    toc: false
    number-sections: true
    colorlinks: true
    link-citations: true
    prompt: false
    include-in-header:
      text: |
        \usepackage{common}
editor_options: 
  chunk_output_type: console
pdf-engine: pdflatex
geometry:
  - left=0.75in
  - right=0.75in
  - top=0.75in
  - bottom=1.00in
code-block-bg: "#FAFAFA"
code-block-border-left: "#C2E3FF"
bibliography: references.bib
---

```{r}
#| echo: false
#| message: false
set.seed(1234)
library(saevws)

print_interface = function(x, ...)
{
	# 0. Convert `...` to a list. If empty, add a default.
    # 1. Get the interface of the function in a printable form; avoid line
    #    breaks.
    # 2. Replace the token 'function ' with the name of the function.
    # 3. Call str_wrap to handle any wrapping if the line is long.
    # 4. Print the wrapped line and replace newline characters with actual
    #    newlines.

	vargs = list(...)
	if (length(vargs) == 0) { vargs = list(exdent = 4) }
	desc = str(x, give.attr = F, width = Inf, strict.width = "wrap") |>
		capture.output()
	vargs$string = sub(
		pattern = "function ",
		replacement = deparse(substitute(x)),
		x = desc,
		fixed = TRUE)
	do.call(stringr::str_wrap, vargs) |> cat(collapse = "\n")
}
```

# Introduction {#sec-intro}

This document describes replication materials for the article "Self-Tuned
Rejection Sampling within Gibbs and a Case Study in Small Area Estimation". It
includes an R package, code for the two simulations, and two data analysis.

@sec-outline presents an outline of the contents of the repo. @sec-setup
describes setup to run the replication materials. @sec-package gives an
overview of the `saevws` package, including installation and basic usage.
@sec-scripts lists the scripts used for the conditional simulation, posterior
simulation, and data analysis. @sec-session gives a comprehensive list of
packages and versions used in the computational studies, for reproducibility.
@sec-amh provides from details about the adaptive Metropolis-Hastings algorithm
used for comparison in the data analyses.

# Outline {#sec-outline}

Here is an outline of the contents of the `saevws-replication` repo.

- `guide`: contains the present document and the source files used to generate
  it.
- `joint`: materials for the "Joint SAE model" based on @You-2021.
	+ `01-sim-cond`: simulation generating from a single conditional of the SAE
	  model under consideration.
	+ `02-sim-post`: simulation generating from the posterior of the SAE model
	  under consideration.
	+ `03-sae`: data analysis with the SAIPE dataset.
	+ `data`: the SAIPE dataset and codes / inputs to assemble it.
	+ `shared`: defines R functions used in several seconds of 
- `saevws`: an R package with the Gibbs samplers and supporting functions.
  See @sec-package of the present document.
- `unmatch`: materials for the "Unmatched SAE" model on @You-Rao-2002.
	+ `01-generated`: analysis based on the generated dataset.
	+ `data`: a generated dataset based on a model in @You-Rao-2002.

# Setup {#sec-setup}

A working installation of R [@Rcore-2026] is required with the following
packages (and their dependencies).

- `coda` [@coda]
- `jsonlite` [@jsonlite]
- `knitr` [@knitr]
- `mcmcse` [@mcmcse]
- `Rcpp` [@Rcpp],
- `statmod` [@statmod]
- `tidyverse` [@tidyverse],
- `xtable` [@xtable]

The above packages are available on CRAN at the time of this writing.
Additionally, development versions of the `fntl` [@fntl] and `vws` [@vws]
packages are needed. Here we demonstrate installing them using the `pak`
package [@pak].

```{r}
#| eval: false
#| prompt: true
pak::pkg_install("andrewraim/fntl@v0.1.3-saevws")
pak::pkg_install("andrewraim/vws@v0.3.1-saevws")
```

The `saevws` package included in the repo can now be installed. The following
uses the `pak` package to accomplish this. (The placeholder `/path/to/` should
be replaced with the actual path.)

```{r}
#| eval: false
#| prompt: true
pak::pkg_install("/path/to/saevws")
```

# `saevws` R Package {#sec-package}

The `saevws` package exports several functions for use in R. The primary
functions are described here. See the associated manual page for each function
for more information.

The following are for the Gibbs sampler in the Joint SAE example.

::: {.callout-note appearance="minimal"}
```{r}
#| echo: false
print_interface(gibbs_joint)
print_interface(control_joint)
print_interface(init_joint)
print_interface(fixed_joint)
```
:::

The following are for the Gibbs sampler in the Unmatched SAE example.

::: {.callout-note appearance="minimal"}
```{r}
#| echo: false
print_interface(gibbs_unmatch)
print_interface(control_unmatch)
print_interface(init_unmatch)
print_interface(fixed_unmatch)
```
:::

The following is used to specify the within-Gibbs sampling method in both
examples.

::: {.callout-note appearance="minimal"}
```{r}
#| echo: false
print_interface(control_inner)
```
:::

These functions are used for plotting.

::: {.callout-note appearance="minimal"}
```{r}
#| echo: false
print_interface(traceplots)
print_interface(plot_comps)
print_interface(plot_rejects)
print_interface(plot_tunes)
```
:::

# Scripts to Reproduce Results in the Article {#sec-scripts}

The following codes in `joint/01-sim-cond` produce the conditional simulation study in Section 6.1 of the article.

- `imh.R`: Simulations with IMH.
- `vws.R`: Simulations with VWS.

The following codes in `joint/02-sim-post` produce the posterior simulation study in Section 6.2 of the article.

- `gen.R`: Generate folder structure for the study. Each level of the study is
  placed into a separate folder.
- `sim.R`: Run one level of the study.
- `analyze.R`: Analyze the results from the completed study and produce 
  summaries.

The `joint/03-sae` folder contains a single script `saipe.R` for the analysis in Section 7 of the article.

The `joint/data` folder contains the SAIPE dataset and scripts and inputs to
construct it.

- `saipe.csv`: the prepared dataset.
- `cntysnap.csv`: Supplemental Nutrition Assistance Program (SNAP) data for
  2022, to be used as a covariate with 2023 SAIPE data.
- `co-est2023-alldata.csv`: 2023 population estimates from the U.S. Census
  Bureau's Population Estimates Program. The data may be obtained from the
  Census Bureau [website][pep].
- `build-data.R`: Code to construct the dataset `saipe.csv`.

The `unmatch/01-generated` folder contains a single `unmatch.R` for the analysis in the supplement.

The `unmatch/data` folder contains the (simulated) dataset for the Unmatched
SAE model, along with scripts and inputs to construct it.

- `build-data.R`: Code to construct the dataset `unmatch.csv`.
- `unmatch.csv`: the prepared dataset.
- `mu-true.csv`: the values of the latent variable used to generate the data.

[pep]: https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-total.html

# Session Information {#sec-session}

Results in the paper were produced in an R environment with the following
packages and versions.

```{r}
#| eval: false
#| message: false
#| warning: false
library(coda)
library(jsonlite)
library(knitr)
library(mcmcse)
library(Rcpp)
library(statmod)
library(tidyverse)
library(xtable)
library(vws)
```

```{r}
sessionInfo()
```

# Adaptive Metropolis-Hastings Algorithm {#sec-amh}

An Adaptive Metropolis-Hastings (AMH) method is implemented in our studies for comparison. The particular variant ("SCAM") is from Section 3.1 of @Roberts-Rosenthal-2009 which is based on @Haario-Saksman-Tamminen-2005. The objective is to draw $\vec{X} = (X_1, \ldots, X_k)$ from target $\pi$. A chain $\vec{x}^{(r)}$, $r = 0, 1, 2, \ldots$, produced by an MCMC method can approximate such draws. Using a Metropolis-within-Gibbs approach to draw component $x_j$, an "optimal" proposal distribution for Metropolis sampling would be $X_j^* \sim \text{N}\big( x_j^{(r)}, (2.38)^2 \cdot \text{Var}_{\pi}(X_j) \big)$. We can then accept $x_j^{(r+1)} = x_j^*$ with probability
<!-- -->
\begin{align*}
\alpha(x_j^*, x_j^{(r)} \mid x_1, \ldots x_{j-1}, x_{j+1}, \ldots, x_k) =
\min\left\{ 1,
\frac {
\pi(x_1, \ldots, x_{j-1}, x_j^*, x_{j+1}, \ldots, x_k)
}{
\pi(x_1, \ldots, x_{j-1}, x_j^{(r)}, x_{j+1}, \ldots, x_k)
}
\right\},
\end{align*}
<!-- -->
and let $x_j^{(r+1)} = x_j^{(r)}$ with the remaining probability. The constant $2.38$ arises in MCMC theory. The variance $\text{Var}_{\pi}(X_j)$ is computed under the target and therefore is generally not accessible except through approximation by an MCMC method. The objective of SCAM is to build up an approximation to $\text{Var}_{\pi}(X_j)$ within the MCMC. @Roberts-Rosenthal-2009 use SCAM with the proposal $\text{N}(x_j^{(r)}, v_j^{(r)})$, where
<!-- -->
\begin{align*}
v_j^{(r)} =
\begin{cases}
v_j^{(0)}, & \text{$r \leq r_0$}, \\
s^2 \{ \tilde{\text{V}}_j^{(r)} + \epsilon \}, & \text{$r > r_0$}, \\
\end{cases}
\end{align*}
<!-- -->
with $\epsilon$ chosen to be a small positive number, $s = 2.4$ as a rounded version of the 2.38 optimality constant, initial value $v_j^{(0)}$, and empirical variance
<!-- -->
\begin{align*}
\tilde{\text{V}}_j^{(r)} =
\frac{1}{r-1} \sum_{i=0}^{r-1} (x_j^{(i)} - \bar{x}_j^{(r)})^2,
\end{align*}
<!-- -->
with associated mean
<!-- -->
\begin{math}
\bar{x}_j^{(r)} = \frac{1}{r} \sum_{i=0}^{r-1} x_j^{(i)}.
\end{math}
<!-- -->
Once the element $x_j^{(r)}$ of the chain is determined at the end of step $r$, the empirical mean and variance can be updated using the recursions so that it is not necessary to retain all of the draws:
<!-- -->
\begin{align*}
\bar{x}_j^{(r+1)} =
\begin{cases}
x_j^{(0)}, & \text{$r = 0$}, \\
\frac{1}{r+1} \{r \bar{x}_j^{(r)} + x_j^{(r)} \}, & \text{$r > 0$}, \\
\end{cases}
\end{align*}
<!-- -->
and
<!-- -->
\begin{align*}
\tilde{\text{V}}_j^{(r+1)} =
\begin{cases}
\{ x_j^{(0)} - x_j^{(1)} \}^2 / 2, & \text{$r = 1$}, \\
\frac{r-1}{r} \tilde{\text{V}}_j^{(r)}
+ \{ \bar{x}_j^{(r)} \}^2
+ \frac{1}{r} \{ x_j^{(r+1)} \}^2
- \frac{r+1}{r} \{ \bar{x}_j^{(r+1)} \}^2, & \text{$r > 1$}, \\
\end{cases}
\end{align*}

For the joint SAE application [@You-2021], we use an exponential transformation to generate positive proposal values from the normal proposal distribution for $\sigma_i^2$. This results in the inclusion of a Jacobian term in the Metropolis ratio as follows. Let $\phi^*$ be a draw from proposal $\text{N}(\phi^{(r)}, v_i^{(r)})$ with $\phi^{(r)} = \log \sigma_i^{2{(r)}}$. The density with respect to $\phi^*$ (with proportionality constant omitted) can be obtained by transformation as
<!-- -->
\begin{align*}
f(\exp(\phi^*) \mid \rest) &\propto
f_\text{IG}(\exp(\phi^*) \mid \kappa_i, \lambda_i)
f_\text{LN}(\exp(\phi^*) \mid \mu_i, \tau^2)
\exp(\phi^*),
\end{align*}
<!-- -->
where the term $e^\phi = \frac{d}{d \phi} e^\phi$ is from the Jacobian of the transformation from $\sigma_i^2$ to $\phi$. The acceptance ratio for Metropolis-Hastings is then
<!-- -->
\begin{align}
\varpi_i^{(r)}
&= \min\left\{ 1, \frac{
f_\text{IG}(\exp(\phi^*) \mid \kappa_i, \lambda_i) \cdot
f_\text{LN}(\exp(\phi^*) \mid \mu_i, \tau^2) \cdot
\exp(\phi^*)
}{
f_\text{IG}(\exp(\phi^{(r)}) \mid \kappa_i, \lambda_i) \cdot
f_\text{LN}(\exp(\phi^{(r)}) \mid \mu_i, \tau^2) \cdot
\exp(\phi^{(r)})
} \right\}.
\label{eqn:joint-ratio}
\end{align}
<!-- -->
For the unmatched SAE application [@You-Rao-2002], we draw proposal variate $\mu_i^* \sim \text{N}(\mu^{(r)}, v_i^{(r)})$ and accept with probability
<!-- -->
\begin{align}
\varpi_i^{(r)}
&= \min\left\{ 1, \frac{
f_\text{LN}(\mu_i^* \mid \vec{x}_i^\top \vec{\beta}, \tau^2) \cdot
f_\text{N}(\mu_i^* \mid y_i, \sigma_i^2)
}{
f_\text{LN}(\mu^{(r)} \mid \vec{x}_i^\top \vec{\beta}, \tau^2) \cdot
f_\text{N}(\mu^{(r)} \mid y_i, \sigma_i^2)
} \right\}.
\label{eqn:unmatch-ratio}
\end{align}
<!-- -->
Note that the Gaussian proposal cancels from both \eqref{eqn:joint-ratio} and \eqref{eqn:unmatch-ratio} because of symmetry. 

# References {-}
