---
title: |
  Self-Tuned Rejection Sampling within Gibbs and a Case Study in Small Area
  Estimation
subtitle: Guide to Replication Materials
author: Andrew M. Raim, Kyle M. Irimata, and James A. Livsey
date: now
date-format: "YYYY-MM-DD"
thanks: |
  `andrew.raim@census.gov, kyle.m.irimata@census.gov, jamesliv@buffalo.edu`.
  Document compiled on `{r} format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.
format:
  pdf:
    fontsize: 10pt
    indent: false
    toc: false
    number-sections: true
    colorlinks: true
    link-citations: true
    prompt: false
    include-in-header:
      text: |
        \usepackage{common}
editor_options: 
  chunk_output_type: console
pdf-engine: pdflatex
geometry:
  - left=0.75in
  - right=0.75in
  - top=0.75in
  - bottom=1.00in
code-block-bg: "#FAFAFA"
code-block-border-left: "#C2E3FF"
bibliography: references.bib
---

```{r}
#| echo: false
#| message: false
set.seed(1234)
library(saevws)
printfn = function(x) {
	cat(x, "=\n")
	str(get(x), give.attr = FALSE, width = 80)
}

print_interface = function(x, ...)
{
	# 0. Convert `...` to a list. If empty, add a default.
    # 1. Get the interface of the function in a printable form; avoid line
    #    breaks.
    # 2. Replace the token 'function ' with the name of the function.
    # 3. Call str_wrap to handle any wrapping if the line is long.
    # 4. Print the wrapped line and replace newline characters with actual
    #    newlines.

	vargs = list(...)
	if (length(vargs) == 0) { vargs = list(exdent = 4) }
	desc = str(x, give.attr = F, width = Inf, strict.width = "wrap") |>
		capture.output()
	vargs$string = sub(
		pattern = "function ",
		replacement = deparse(substitute(x)),
		x = desc,
		fixed = TRUE)
	do.call(stringr::str_wrap, vargs) |> cat(collapse = "\n")
}
```

# Introduction

This document describes replication materials for the article "Self-Tuned
Rejection Sampling within Gibbs and a Case Study in Small Area Estimation". It
includes an R package, code for two simulations, and a data analysis.
The following folders are included in the materials.

- `01-sim-cond`: simulation generating from a single conditional of the SAE
  model under consideration.
- `02-sim-post`: simulation generating from the posterior of the SAE model under
  consideration.
- `03-sae`: data analysis with the SAIPE dataset.
- `data`: the SAIPE dataset and codes /inputs to assemble it.
- `vws`: An R package that supports construction of VWS samplers.
- `saevws`: an R package for the Gibbs sampler and supporting functions. It can
  be configured with basic vertical weighted strips (VWS), self-tuned VWS,
  and independent Metropolis-Hastings (IMH) for the
  $\sigma_1^2, \ldots, \sigma_m^2$ steps.

A working installation of R [@Rcore2026] is required with the following
packages (and their dependencies). These packages are available on CRAN at the
time of this writing.

- `devtools` [@devtools],
- `fntl` [@fntl]
- `jsonlite` [@jsonlite]
- `knitr` [@knitr]
- `mcmcse` [@mcmcse]
- `Rcpp` [@Rcpp],
- `statmod` [@statmod]
- `tidyverse` [@tidyverse],
- `xtable` [@xtable]

Additionally, the included `vws` package must be installed.

```{r}
#| eval: false
#| prompt: true
devtools::install_local("/path/to/vws")
```

@sec-package gives an overview of the `saevws` package, including installation and basic usage. @sec-scripts lists the scripts used for the conditional simulation, posterior simulation, and data analysis. @sec-session gives a comprehensive list of packages and versions used in the computational studies, for reproducibility.

# `saevws` R Package {#sec-package}

The `saevws` package can be installed using an R command like the following.

```{r}
#| eval: false
#| prompt: true
devtools::install_local("/path/to/saevws")
```

**TBD: Needs update**
The package exports several functions for use in R:

- `gibbs_joint` invokes the Gibbs sampler.
- `control_joint` constructs a control object with additional arguments for
  the Gibbs sampler.
- `get_vws_control` constructs a control object with additional arguments for
  VWS within the Gibbs sampler.
- `get_init` constructs an object with initial values for the Gibbs sampler.
- `get_fixed` constructs an object to indicate that certain elements of the
  MCMC chain are to remain fixed in the Gibbs sampler. This is primarily used
  for testing the code.
- `traceplots` is a utility to make trace plots from a matrix with draws.

These functions' interfaces are shown below; see the associated manual pages
for details about arguments, outputs, etc..

```{r}
#| echo: false
print_interface(gibbs_joint)
print_interface(control_joint)
print_interface(init_joint)
print_interface(fixed_joint)

print_interface(gibbs_unmatch)
print_interface(control_unmatch)
print_interface(init_unmatch)
print_interface(fixed_unmatch)

print_interface(control_inner)
print_interface(traceplots)
```

# Scripts {#sec-scripts}

The following codes in `01-sim-cond` produce the conditional simulation study in Section 5.1 of the article.

- `metro.R`: Simulations with IMH.
- `vws.R`: Simulations with VWS.
- `samplers.cpp`: C++ functions for samplers consisting of one IMH step and one
  VWS step.

The following codes in `02-sim-post` produce the posterior simulation study in Section 5.2 of the article.

- `gen.R`: Generate folder structure for the study. Each level of the study is
  placed into a separate folder.
- `sim.R`: Run one level of the study.
- `analyze.R`: Analyze the results from the completed study and produce 
  summaries.

The `03-sae` folder contains a single script `saipe.R` that produces the data analysis in Section 6 of the article.

The `data` folder contains scripts and inputs to create the example SAIPE
dataset.

- `saipe.csv`: the prepared dataset.
- `cntysnap.csv`: Supplemental Nutrition Assistance Program (SNAP) data for
  2022, to be used as a covariate with 2023 SAIPE data.
- `co-est2023-alldata.csv`: 2023 population estimates from the U.S. Census
  Bureau's Population Estimates Program. The data may be obtained from the
  Census Bureau [website][pep].
- `build-data.R`: Code to construct the dataset `saipe.csv`.

[pep]: https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-total.html

# Session Information {#sec-session}

Results in the paper were produced in an R environment with the following
packages and versions.

```{r}
#| eval: false
#| message: false
#| warning: false
library(devtools)
library(tidyverse)
library(Rcpp)
library(knitr)
library(mcmcse)
library(jsonlite)
library(xtable)
library(vws)
library(statmod)
```

```{r}
sessionInfo()
```

# Adaptive Metropolis-Hastings Algorithm {#sec-am}

An Adaptive Metropolis-Hastings (AMH) method is implemented in our studies for comparison. The particular variant ("SCAM") is from Section~3.1 of @Roberts-Rosenthal-2009 which is based on @Haario-Saksman-Tamminen-2005. The objective is to draw $\vec{X} = (X_1, \ldots, X_k)$ from target $\pi$. A chain $\vec{x}^{(r)}$, $r = 0, 1, 2, \ldots$, produced by an MCMC method can approximate such draws. Using a Metropolis-within-Gibbs approach to draw component $x_j$, an "optimal" proposal distribution for Metropolis sampling would be $X_j^* \sim \text{N}(x_j^{(r)}, (2.38)^2 \text{Var}_{\pi}(X_j))$. We can then accept $x_j^{(r+1)} = x_j^*$ with probability
<!-- -->
\begin{align*}
\alpha(x_j^*, x_j^{(r)}, \vec{x}_{\setminus j}) =
\min\left\{ 1,
\frac {
\pi(x_1, \ldots, x_{j-1}, x_j^*, x_{j+1}, \ldots, x_k)
}{
\pi(x_1, \ldots, x_{j-1}, x_j^{(r)}, x_{j+1}, \ldots, x_k)
}
\right\},
\end{align*}
<!-- -->
where $\vec{x}_{\setminus j}$ denotes the last draws of components $\ell \neq j$ from the chain, and let $x_j^{(r+1)} = x_j^{(r)}$ with the remaining probability. The constant $2.38$ arises in MCMC theory. The variance $\text{Var}_{\pi}(X_j)$ is computed under the target and therefore is generally not accessible except through approximation by an MCMC method. The objective of SCAM is to build up an approximation to $\text{Var}_{\pi}(X_j)$ within the MCMC. @Roberts-Rosenthal-2009 use SCAM with the proposal $\text{N}(x_j^{(r)}, v_j^{(r)})$, where
<!-- -->
\begin{align*}
v_j^{(r)} =
\begin{cases}
v_j^{(0)}, & \text{$r \leq r_0$}, \\
s^2 \{ \tilde{\text{V}}_j^{(r)} + \epsilon \}, & \text{$r > r_0$}, \\
\end{cases}
\end{align*}
<!-- -->
with $\epsilon$ chosen to be a small positive number, $s = 2.4$ as a rounded version of the 2.38 optimality constant, initial value $v_j^{(0)}$, and empirical variance
<!-- -->
\begin{align*}
\tilde{\text{V}}_j^{(r)} =
\frac{1}{r-1} \sum_{i=0}^{r-1} (x_j^{(i)} - \bar{x}_j^{(r)})^2,
\end{align*}
<!-- -->
with associated mean
<!-- -->
\begin{math}
\bar{x}_j^{(r)} = \frac{1}{r} \sum_{i=0}^{r-1} x_j^{(i)}.
\end{math}
<!-- -->
Once the element $x_j^{(r)}$ of the chain is determined at the end of step $r$, the empirical mean and variance can be updated using the recursions so that it is not necessary to retain all of the draws:
<!-- -->
\begin{align*}
\bar{x}_j^{(r+1)} =
\begin{cases}
x_j^{(0)}, & \text{$r = 0$}, \\
\frac{1}{r+1} \{r \bar{x}_j^{(r)} + x_j^{(r)} \}, & \text{$r > 0$}, \\
\end{cases}
\end{align*}
<!-- -->
and
<!-- -->
\begin{align*}
\tilde{\text{V}}_j^{(r+1)} =
\begin{cases}
\{ x_j^{(0)} - x_j^{(1)} \}^2 / 2, & \text{$r = 1$}, \\
\frac{r-1}{r} \tilde{\text{V}}_j^{(r)}
+ \{ \bar{x}_j^{(r)} \}^2
+ \frac{1}{r} \{ x_j^{(r+1)} \}^2
- \frac{r+1}{r} \{ \bar{x}_j^{(r+1)} \}^2, & \text{$r > 1$}, \\
\end{cases}
\end{align*}

For the joint SAE application [@You2021], we use an exponential transformation to generate positive proposal values from the normal proposal distribution for $\sigma_i^2$. This results in the inclusion of a Jacobian term in the Metropolis ratio as follows. Let $\phi^*$ be a draw from proposal $\text{N}(\phi^{(r)}, v_i^{(r)})$ with $\phi^{(r)} = \log \sigma_i^{2{(r)}}$. The density with respect to $\phi^*$ (with proportionality constant omitted) can be obtained by transformation as
<!-- -->
\begin{align*}
f(\exp(\phi^*) \mid \rest) &\propto
f_\text{IG}(\exp(\phi^*) \mid \kappa_i, \lambda_i)
f_\text{LN}(\exp(\phi^*) \mid \mu_i, \tau^2)
\exp(\phi^*),
\end{align*}
<!-- -->
where the term $e^\phi = \frac{d}{d \phi} e^\phi$ is from the Jacobian of the transformation from $\sigma_i^2$ to $\phi$. The acceptance ratio for Metropolis-Hastings is then
<!-- -->
\begin{align}
\varpi_i^{(r)}
&= \min\left\{ 1, \frac{
f_\text{IG}(\exp(\phi^*) \mid \kappa_i, \lambda_i) \cdot
f_\text{LN}(\exp(\phi^*) \mid \mu_i, \tau^2) \cdot
\exp(\phi^*)
}{
f_\text{IG}(\exp(\phi^{(r)}) \mid \kappa_i, \lambda_i) \cdot
f_\text{LN}(\exp(\phi^{(r)}) \mid \mu_i, \tau^2) \cdot
\exp(\phi^{(r)})
} \right\}.
\label{eqn:joint-ratio}
\end{align}
<!-- -->
For the unmatched SAE application [@You-Rao-2002], we draw proposal variate $\mu_i^* \sim \text{N}(\mu^{(r)}, v_i^{(r)})$ and accept with probability
<!-- -->
\begin{align}
\varpi_i^{(r)}
&= \min\left\{ 1, \frac{
f_\text{LN}(\mu_i^* \mid \vec{x}_i^\top \vec{\beta}, \tau^2) \cdot
f_\text{N}(\mu_i^* \mid y_i, \sigma_i^2)
}{
f_\text{LN}(\mu^{(r)} \mid \vec{x}_i^\top \vec{\beta}, \tau^2) \cdot
f_\text{N}(\mu^{(r)} \mid y_i, \sigma_i^2)
} \right\}.
\label{eqn:unmatch-ratio}
\end{align}
<!-- -->
Note that the Gaussian proposal cancels from both \eqref{eqn:joint-ratio} and \eqref{eqn:unmatch-ratio} because of symmetry. 

# References {-}
